---
title: "Script for Downloading Qualitative Data from Webpages"
author: "Adrien Ratsimbaharison"
date: "9/18/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


*The following procedures are based on the instructions provided by Bradley Boehmke in [Scraping HTML Text](http://bradleyboehmke.github.io/2015/12/scraping-html-text.html):*


To extract text from a webpage of interest, we specify what HTML elements we want to select by using `html_nodes()`. For instance, if we want to scrape the primary heading for the Web Scraping Wikipedia webpage we simply identify the `h1 node` as the node we want to select. `html_nodes()` will identify all `h1 nodes` on the webpage and return the HTML element. In our example we see there is only one `h1` node on this webpage.  

```{r}
library(rvest)

scraping_wiki <- read_html("https://en.wikipedia.org/wiki/Web_scraping")

scraping_wiki %>%
        html_nodes("h1")
```

To extract only the heading text for this `h1 node`, and not include all the `HTML syntax` we use `html_text()` which returns the heading text we see at the top of the Web Scraping Wikipedia page.

```{r}
scraping_wiki %>%
        html_nodes("h1") %>%
        html_text()
```

To extract much of the text on this webpage which is in paragraph form. We can follow the same process illustrated above but instead we’ll select all `p nodes`. This selects the 17 paragraph elements from the web page; which we can examine by subsetting the list `p_nodes` to see the first line of each paragraph along with the HTML syntax. Just as before, to extract the text from these nodes and coerce them to a character string we simply apply `html_text()`.

```{r}
p_text <- scraping_wiki %>%
        html_nodes("p") %>%
        html_text()

p_text[1]
```

### Example with the wikipedia page on the second civil war in Sudan

```{r}
library(rvest)

Url <- "Source: https://en.wikipedia.org/wiki/Second_Sudanese_Civil_War"

web_page <- read_html("https://en.wikipedia.org/wiki/Second_Sudanese_Civil_War")
# extracting the header
header <- web_page %>%
        html_nodes("h1") %>%
        html_text()

# extracting the text
text <- web_page %>%
        html_nodes("p") %>%
        html_text()

# saving the text to a local repository

Sudan2 <- writeLines(c(header, Url, text), "qualitativeData/Sudan2.txt")

```

