---
title: "Zimbabwe"
author: ""
date: ""
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GNI per capita and development level

The data set shows the GNI per capita and development level of the country each year.


```{r GNI per capita data, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(readr)
library(dplyr)

developmentData <- read_csv("WGI2popDevIneqPovRegimeConflict1V2.csv")

developmentVariables <- c("country", "iso3c", "date","GNIperCapita","devCategory")
developmentData <- developmentData[developmentVariables]


# data by country
# country = x
x <- developmentData[developmentData$country == "Zimbabwe",]
x

```



## GDP growth and economic growth classification

The data set shows the GDP growth and the ecomic growth classification of the country each year.


```{r GDP growth data, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}

library(readr)
library(dplyr)

GDPgrowthData <- read_csv("WGI2popDevIneqPovRegimeConflict1V2.csv")

# addind a category of GDP growth
# -Inf to 0: negative growth
# 0 to 2.4: slow growth
# 2.5 to 5.4: moderate growth
# 5.5 and more: fast growth

GDPgrowthData <- GDPgrowthData %>% mutate(GDPgrowthCategory4=cut(GDPannualGrowthRate, breaks=c(-Inf, 0, 2.5, 5.5, Inf), labels=c("negative grwoth","slow growth","moderate growth", "fast growth")))

GDPgrowthDataVariables <- c("country", "iso3c", "date", "GDPannualGrowthRate", "GDPgrowthCategory4")
GDPgrowthData <- GDPgrowthData[GDPgrowthDataVariables]


# data by country
# country = x
x1 <- GDPgrowthData[GDPgrowthData$country == "Zimbabwe",]
x1

```


## Development trend

The plot shows the development trend of the country. 


```{r GNI per capita Plot, echo=FALSE, fig.cap="GNI Per Capita Trend", message=FALSE, warning=FALSE, paged.print=FALSE}
# Development category
# 0-1025: Low Income
# 1026-4035: Lower Middle Income
# 4036-12475:Upper Middle Income
# 12476 and more: High Income

library(ggplot2)
developmentTrend <- ggplot(data=x, aes(x=date, y=GNIperCapita)) +
  geom_line()+
  geom_point() +
  geom_hline(yintercept=1025, linetype="dashed", color = "red", size = 1) +
  geom_hline(yintercept=4035, linetype="dashed", color = "red", size = 1) +
  geom_hline(yintercept=12476, color = "red", size = 1.25) +
  ylim(0, 14000) + 
  xlab("Year") +
  ylab("GNI per capita")

plot(developmentTrend)

# To save the figure:
# png("developmentTrend.png")
# 
# plot(developmentTrend)
# 
# dev.off()

```


## Growth trend

The plot shows the economic growth trend of the country. 


```{r growth trend, echo=FALSE, fig.cap="Economic Growth Trend", message=FALSE, warning=FALSE, paged.print=FALSE}
# Development category
# -Inf to 0: negative growth
# 0 to 2.4: slow growth
# 2.5 to 5.4: moderate growth
# 5.5 and more: fast growth

library(ggplot2)
growthTrend <- ggplot(data=x1, aes(x=date, y=GDPannualGrowthRate)) +
  geom_line()+
  geom_point() +
  geom_hline(yintercept=2.5, linetype="dashed", color = "red", size = 1) +
  geom_hline(yintercept=5.5, linetype="dashed", color = "red", size = 1) +
  geom_hline(yintercept=0, color = "red", size = 1.25) +
  ylim(0, 20) + 
  xlab("Year") +
  ylab("GNI per capita")

plot(growthTrend)

# To save the figure:
# png("growthTrend.png")
# 
# plot(growthTrend)
# 
# dev.off()

```

