---
title: "Script for the Machine Learning Predictions"
author: "Adrien M. Ratsimbaharison"
date: "March 17, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## The six steps in the machine learning predictions:

These steps generally follow the steps suggested by different machine learning practitioners and instructors, including Lantz (2015), Pierobon (2018), and Leek, Peng, & Caffo (2018).

1) Defining the problem 
2) Exploring the Data 
3) Preparing the data: data randomization and partition
4) Modeling and predicting with classification
5) Modeling and predicting with regression
6) Conclusion

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
#  Loading the required packages:
library(caret)
library(dplyr)
library(ggplot2)
library(readr)
library(rattle)

```


### Step 1: Defining the problem

When performing machine learning techniques, it is always a good idea to define exactly What you are trying to predict. In our case, we are trying to predict:

1) the classification of countries either as "stable" (with stability scores greater than 0) or as "unstable" (with stability scores less than 0), and
2) the stability scores of countries based on some predictors.
  
In terms of machine learning, we are dealing in this case with a multivariate supervised machine learning problem in which we have to predict a binary and multi-class outcome (thus we will use classification techniques) and a numeric outcome (thus we will use a mutliple regression technique).  


### Step 2: Reading and exploring the Data in r

Our data were downloaded from different sources and merged in a single data frame (See script on downloading quantitative data for the information on how the data were collected and manipulated).

We just need here to read the data into r and explore the features.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# reading the data into r

WGIdevRegimeType <- read.csv("WGIdevRegimeType.csv")

# looking at the different types of variables and correcting their classes

WGIdevRegimeType$stabilityDummy <- as.factor(WGIdevRegimeType$stabilityDummy)

glimpse(WGIdevRegimeType)

```

Before preparing the data for the modeling, we need to take a look at the target variables: stabilityDummy and stability.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# summary statistics of the target variables
table(WGIdevRegimeType$stabilityDummy)

summary(WGIdevRegimeType$stability)

# plotting the target variables

barplot(table(WGIdevRegimeType$stabilityDummy))

hist(WGIdevRegimeType$stability)

```

We need also to take a look at the summary statistics of the other variables of interest, paying attention to the existence of NAs, outliers, and zero variances.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

WGIdevRegimeTypeSummary <- select(WGIdevRegimeType, stability, stabilityDummy, stabilityCategory, corruptionControl, governmentEffectiveness, regulatoryQuality, ruleOfLaw, voiceAndAccountability, GNIperCapita, devCategory, GDPannualGrowthRate, HDI, GINI, povertyHeadCount, polityScore, polityCategory, politicalChange, democ, autoc, durable, xrreg, xrcomp, xropen, xconst, parreg, parcomp, exrec, exconst, polcomp, pr, cl, sum, mean, status, inverse_pr, inverse_cl, inverse_mean, politicalChangeFH, region, subregion)

summary(WGIdevRegimeTypeSummary)

```

From the summary statistics of the other variables, we can see that, among the Polity IV variables, there are some extreme values of -66 (for interruption periods), -77 (for interregnum), and -88 (for transition periods), which are known as "standardized authority codes" (See Marshall, M. G., Gurr, T. R., & Jaggers, K., 2018). Since there is no commonly agreed procedure on how to handle these extreme values among scholars (See Plümper, T., & Neumayer, E., 2010), we decided to remove them from our data, because they may falsify our results.

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Removing the interruption periods (-66), interregnum periods (-77), and transitional periods (-88)

WGIdevRegimeType <- filter(WGIdevRegimeType, democ != -88 & democ != -77 &  democ != -66)

summary(WGIdevRegimeType)
dim(WGIdevRegimeType)

```


### Step 3: Preparing the data: data randomization and feature selection

The data were arranged in a country-year format. Therefore, it is a good idea to randoming the rows before partitioning the dataset

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Randomizing the dataset
set.seed(112)
rows <- sample(nrow(WGIdevRegimeType))
WGIdevRegimeTypeRand <- WGIdevRegimeType[rows, ]



```


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Selecting the variables to be included in the classification model. We need to remove the political stability score (stability) and stability category (stabilityCategory) from this dataset.

WGIdevRegimeTypeRandClass <- select(WGIdevRegimeTypeRand, stabilityDummy, corruptionControl, governmentEffectiveness, regulatoryQuality, ruleOfLaw, voiceAndAccountability, GNIperCapita, GDPannualGrowthRate, HDI, GINI, povertyHeadCount, polityScore, polityCategory, politicalChange, democ, autoc, durable, xrreg, xrcomp, xropen, xconst, parreg, parcomp, exrec, exconst, polcomp, pr, cl, sum, mean, status, inverse_pr, inverse_cl, inverse_mean, politicalChangeFH, region, subregion)

```


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Selecting the variables to be included in the regression model. We need to remove the stability dummy (stabilityDummy) and stability category (stabilityCategory) from this dataset.

WGIdevRegimeTypeRandReg <- select(WGIdevRegimeTypeRand, stability, corruptionControl, governmentEffectiveness, regulatoryQuality, ruleOfLaw, voiceAndAccountability, GNIperCapita, GDPannualGrowthRate, HDI, GINI, povertyHeadCount, polityScore, polityCategory, politicalChange, democ, autoc, durable, xrreg, xrcomp, xropen, xconst, parreg, parcomp, exrec, exconst, polcomp, pr, cl, sum, mean, status, inverse_pr, inverse_cl, inverse_mean, politicalChangeFH, region, subregion)

```


### Step 4: Modeling and predicting with classification


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Partitionning the dataset into training set and testing set for the classification model
set.seed(123)
inTrain1 <- createDataPartition(y = WGIdevRegimeTypeRandClass$stabilityDummy, p = .75, list = FALSE)
training1 <- WGIdevRegimeTypeRandClass[inTrain1,]
testing1 <- WGIdevRegimeTypeRandClass[-inTrain1,]


```


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Training the classification model

classModel <- train(stabilityDummy ~ ., method = "rpart", data = training1,
                    trControl = trainControl(method = "cv", number = 5,
                                           verboseIter = TRUE),
                    preProcess = c("zv", "nzv", "center", "scale")
                    )

print(classModel$finalModel)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Plotting the classification model
fancyRpartPlot(classModel$finalModel)

```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Making prediction and assessing the model performance

classPred <- predict(classModel, testing1)

confusionMatrix(classPred, testing1$stabilityDummy)

```


### Step 5: Modeling and predicting with regression


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Partitionning the dataset into training set and testing set for the regression model
set.seed(234)
inTrain2 <- createDataPartition(y = WGIdevRegimeTypeRandReg$stability, p = .75, list = FALSE)
training2 <- WGIdevRegimeTypeRandReg[inTrain2,]
testing2 <- WGIdevRegimeTypeRandReg[-inTrain2,]
dim(training2)
dim(testing2)

```


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Training the regression model

regModel <- train(stability ~ ., method = "glm", data = training2,
                  trControl = trainControl(method = "cv", number = 5,
                                           verboseIter = TRUE),
                  preProcess = c("zv", "nzv", "center", "scale", "pca")
                  )
regFinalModel <- regModel$finalModel
  
print(regModel)
```

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Regression model diagnostics

plot(regFinalModel, 1, pch = 19)



```


```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# Making prediction and assessing the model performance

regPred <- predict(regModel, testing2)

postResample(pred = regPred, obs = testing2$stability)

```


### Step 6: Conclusion


