---
title: "Script for Comparing Stability Score with Regime Durability and State Fragility"
author: "Adrien Ratsimbaharison"
date: "July 17, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Purpose

The purpose of this script is to compile and compare the other measures state or regime characteristics to see if they are correlated and can be used interchangeably.   

## Loading the packages and data

```{r Required R Packages, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
library(dplyr)
library(ggplot2)
library(readxl)
library(lubridate)
library(psych)
library(Hmisc)
library(PerformanceAnalytics)


```

The data on stability score and durability from the full dataset:

```{r data on stability and durability, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Loading the data on stability and durability from the full dataset (WGI2popDevIneqPovRegimeConflict1.csv) available in the local repository which include the Polity IV data with regime durability

stabilityDurabilityDataset <- read.csv("WGI2popDevIneqPovRegimeConflict1.csv", header = TRUE)
stabilityDurabilityDataset <- as.data.frame(stabilityDurabilityDataset)

# Selecting the variables of interest
stabilityDurabilityDataset <- select(stabilityDurabilityDataset, country, date, stability, durability = durable)

# Correcting the variable types
stabilityDurabilityDataset$country <- as.character(stabilityDurabilityDataset$country)
stabilityDurabilityDataset$date <- as.numeric(stabilityDurabilityDataset$date)
stabilityDurabilityDataset$durability <- as.numeric(stabilityDurabilityDataset$durability)

# Filtering the data for the period 2010 to 2017
stabilityDurabilityDataset <- filter(stabilityDurabilityDataset, date > 2009)
str(stabilityDurabilityDataset)

# Downloading the data on state fragility index (SFIv2017) for the Center of Systemic Peace
# url <- "http://www.systemicpeace.org/inscr/SFIv2017.xls"
# destfile <- "SFIv2017.xls"
# curl::curl_download(url, destfile)
# SFIv2017 <- read_excel(destfile)
#
# Making SFIv2017 available in the local repository
# SFIv2017 <- write.csv(SFIv2017, "SFIv2017.csv")

# Loading the data on state fragility index (SFIv2017)
SFIv2017 <- read.csv("SFIv2017.csv", header = TRUE)
SFIv2017 <- as.data.frame(SFIv2017)
# Selecting the variables of interest from SFIv2017
SFIv2017 <- select(SFIv2017, country, date = year, SFI = sfi)
# Filtering the data for the period 2010-2017
SFIv2017 <- filter(SFIv2017, date > 2009)

# Correcting the variable types
SFIv2017$country <- as.character(SFIv2017$country)
SFIv2017$date <- as.numeric(SFIv2017$date)
SFIv2017$SFI <- as.numeric(SFIv2017$SFI)

# Correcting country names
# The following country names have been changed:
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Cote d'Ivoire", "Côte d'Ivoire")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Gambia", "Gambia, The")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Cape Verde", "Cabo Verde")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Congo-Brazzaville", "Congo, Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Dem. Rep. of Congo", "Congo, Dem. Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Timor Leste", "Timor-Leste")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Egypt", "Egypt, Arab Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Syria", "Syrian Arab Republic")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Iran", "Iran, Islamic Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Kyrgyzstan", "Kyrgyz Republic")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Korea, North", "Korea, Dem. People's Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Russia", "Russian Federation")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Taiwan", "Taiwan, China")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Korea South", "Korea, Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Laos", "Lao PDR")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Yemen", "Yemen, Rep.")
SFIv2017$country <- replace(SFIv2017$country, SFIv2017$country == "Venezuela", "Venezuela, RB")

# The following country names have been deleted:
SFIv2017 <- SFIv2017[! SFIv2017$country == "Serbia & Montenegro",]
SFIv2017 <- SFIv2017[! SFIv2017$country == "Sudan (North)",]

# # Downloading the data on fragile state index (fsi-year) from Fund for Peace at: https://fragilestatesindex.org/excel/ for the period 2010-2017
# 
# # For 2017
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2017.xlsx"
# destfile <- "fsi_2017.xlsx"
# curl::curl_download(url, destfile)
# fsi_2017 <- read_excel(destfile)
# str(fsi_2017)
# 
# # For 2016
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2016.xlsx"
# destfile <- "fsi_2016.xlsx"
# curl::curl_download(url, destfile)
# fsi_2016 <- read_excel(destfile)
# str(fsi_2016)
# 
# # For 2015
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2015.xlsx"
# destfile <- "fsi_2015.xlsx"
# curl::curl_download(url, destfile)
# fsi_2015 <- read_excel(destfile)
# str(fsi_2015)
# 
# # For 2014
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2014.xlsx"
# destfile <- "fsi_2014.xlsx"
# curl::curl_download(url, destfile)
# fsi_2014 <- read_excel(destfile)
# str(fsi_2014)
# 
# # For 2013
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2013.xlsx"
# destfile <- "fsi_2013.xlsx"
# curl::curl_download(url, destfile)
# fsi_2013 <- read_excel(destfile)
# str(fsi_2013)
# 
# # For 2012
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2012.xlsx"
# destfile <- "fsi_2012.xlsx"
# curl::curl_download(url, destfile)
# fsi_2012 <- read_excel(destfile)
# str(fsi_2012)
# 
# # For 2011
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2011.xlsx"
# destfile <- "fsi_2011.xlsx"
# curl::curl_download(url, destfile)
# fsi_2011 <- read_excel(destfile)
# str(fsi_2011)
# 
# # For 2010
# url <- "https://fragilestatesindex.org/wp-content/uploads/data/fsi-2010.xlsx"
# destfile <- "fsi_2010.xlsx"
# curl::curl_download(url, destfile)
# fsi_2010 <- read_excel(destfile)
# str(fsi_2010)
# 
# # Merging the data on fragile state index (fsi2010_2017) from Fund for Peace
# fsi2010_2017 <- rbind(fsi_2017, fsi_2016, fsi_2015, fsi_2014, fsi_2013, fsi_2012, fsi_2011, fsi_2010)
# str(fsi2010_2017)
# 
# # Making fsi2010_2017 available in the local repository
# fsi2010_2017 <- write.csv(fsi2010_2017, "fsi2010_2017.csv")
# 

# Loading the data on fragile state index (fsi2010_2017) from 
fsi2010_2017 <- read.csv("fsi2010_2017.csv", header = TRUE)
fsi2010_2017 <- as.data.frame(fsi2010_2017)

# Selecting the variables of interest from SFIv2017
fsi2010_2017 <- select(fsi2010_2017, country = Country, date = Year, fsi = Total)
# Correcting the variable types
fsi2010_2017$country <- as.character(fsi2010_2017$country)
fsi2010_2017$fsi <- as.numeric(fsi2010_2017$fsi)

# Changing the date format
fsi2010_2017$date <- as.Date(fsi2010_2017$date)
fsi2010_2017$date <- format(fsi2010_2017$date,"%Y")
fsi2010_2017$date <- as.numeric(fsi2010_2017$date)
str(fsi2010_2017)

# Correcting country names
# The following country names have been changed:
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Yemen", "Yemen, Rep.")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Cote d'Ivoire", "Côte d'Ivoire")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Gambia", "Gambia, The")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Cape Verde", "Cabo Verde")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Congo Republic", "Congo, Rep.")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Congo Democratic Republic", "Congo, Dem. Rep.")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Egypt", "Egypt, Arab Rep.")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "North Korea", "Korea, Dem. People's Rep.")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Iran", "Iran, Islamic Rep.")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Russia", "Russian Federation")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Venezuela", "Venezuela, RB")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Laos", "Lao PDR")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Syria", "Syrian Arab Republic")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Bahamas", "Bahamas, The")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Macedonia", "Macedonia, FYR")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "Sao Tome and Principe", "São Tomé and Principe")
fsi2010_2017$country <- replace(fsi2010_2017$country, fsi2010_2017$country == "South Korea", "Korea, Rep.")

# Merging the data on stability and durability with SFI and fsi
stabilityDurabilitySFIfsi <- left_join(stabilityDurabilityDataset, SFIv2017, by = c("country", "date")) %>%
  left_join(., fsi2010_2017, by = c("country", "date"))

# Droping NAs
stabilityDurabilitySFIfsi <- na.omit(stabilityDurabilitySFIfsi)
str(stabilityDurabilitySFIfsi)
summary(stabilityDurabilitySFIfsi)

```

## Correlation matrix among the variables

First, we need to generate a correlation matrix with significance:

```{r correlation matrix, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
# selecting the variables to be included in the matrix
stabilityDurabilitySFIfsiScores <- select(stabilityDurabilitySFIfsi, stability, durability, SFI, fsi)
correlationScores <- rcorr(as.matrix(stabilityDurabilitySFIfsiScores))
correlationScores

```


The tables below calculate and plot the correlations among stability, durability, Center of Systemic Peace's SFI and Fund for Peace's fsi:

```{r correlation plots, echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}

# Plot with PerformanceAnalytics
chart.Correlation(stabilityDurabilitySFIfsiScores, histogram=TRUE, pch=19)

# Plot with Hmisc
pairs.panels(stabilityDurabilitySFIfsiScores, scale=TRUE)

```

