---
title: "Understanding the Durability of the Paul Biya Regime in Cameroon: A Micro-Level Analysis
  Using Afrobarometer Data"
author: "Adrien Ratsimbaharison and Benn Bongang"
date: "March 12, 2019"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r the packages, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# loading the required packages
library(haven)
library(survey)
library(dplyr)
library(ggplot2)
library(zeallot)

```

```{r the data, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
# loading the data

# Cameroon Round 5 data (updated July 2015)
library(haven)
# cam_r5_data_fr_july_2015 <- read_sav("http://afrobarometer.org/sites/default/files/data/round-5/cam_r5_data_fr_july_2015.sav")
# write_sav(cam_r5_data_fr_july_2015, "cam_r5_data_fr_july_2015.sav")

cameroonRound5 <- read_sav("cam_r5_data_fr_july_2015.sav")
cameroonRound5 <- as_tibble(cameroonRound5)

# round 6 data
library(haven)
cameroonRound6 <- read_sav("cam_r6_data_2016_eng.sav")
cameroonRound6 <- as_tibble(cameroonRound6)

# round 7 data
library(haven)
cameroonRound7 <- read_sav("cam_r7_data_eng.sav")
cameroonRound7 <- as_tibble(cameroonRound7)

# Selecting the variables of interest for the undertanding the trust in President Paul Biya and the approval of his job performance

cameroonRound7 <- select(cameroonRound7, URBRUR, REGION, EA_SVC_A, EA_SVC_B, EA_SVC_C, EA_SVC_D, EA_FAC_A, EA_FAC_B, EA_FAC_C, EA_FAC_D, EA_FAC_E, EA_FAC_F, EA_FAC_G, EA_SEC_A, EA_SEC_B, EA_SEC_C, EA_SEC_D, EA_SEC_E, EA_ROAD_A, EA_ROAD_B, EA_ROAD_C, Q1, Q2A, Q2AOTHER, Q2B, Q2BOTHER, Q3, Q4A, Q4B, Q5, Q6, Q7, Q8A, Q8B, Q8C, Q8D, Q8E, Q8F, Q9, Q10A, Q10B, Q11A, Q11B, Q12A, Q12B, Q12C, Q12D, Q12E, Q13, Q14, Q15, Q16, Q17, Q18A, Q18B, Q18C, Q18D, Q19A, Q19B, Q19C, Q19D, Q19E, Q20A, Q20B, Q21A, Q21B, Q22, Q23, Q24A, Q24B, Q25A, Q25B, Q25C, Q25D, Q25E, Q25F, Q26A, Q26B, Q26C, Q26D, Q26E, Q27A, Q27B, Q27C, Q28, Q29, Q30, Q31, Q32, Q33, Q34, Q35, Q36, Q37, Q38A, Q38B, Q38C, Q38D, Q38E, Q38F, Q38G, Q39A, Q39B, Q39C, Q40, Q41, Q42A, Q42B, Q42C, Q42D, Q42E, Q42F, Q43A, Q43B, Q43C, Q43D, Q43E, Q43F, Q43G, Q43H, Q43I, Q43J, Q43K, Q43L_CAM, Q43M_CAM, Q44A, Q44B, Q44C, Q44D, Q44E, Q44F, Q44G, Q44H, Q44I, Q44J, Q44K_CAM, Q44L_CAM, Q44M_CAM, Q44N_CAM, Q45, Q46, Q47, Q48A, Q48B, Q48C, Q48D, Q48E, Q48F, Q49A, Q49B, Q49C, Q49D, Q49E, Q49F, Q49G, Q49H, Q49I, Q49J, Q49K, Q49L, Q49M, Q49N, Q49O, Q49P, Q49Q, Q49R, Q49S, Q49T, Q50, Q51, Q52, Q53A, Q53B, Q53C, Q54A, Q54B, Q55PT1, Q55PT1OTHER, Q55PT2, Q55PT2OTHER, Q55PT3, Q55PT3OTHER, Q56A, Q56B, Q56C, Q56D, Q56E, Q56F, Q56G, Q56H, Q56I, Q56J, Q56K, Q56L, Q56M, Q56N, Q56O, Q56P, Q56Q, Q56R, Q56S, Q57A, Q57B, Q57C, Q57D, Q57E, Q57F, Q57G, Q58A, Q58B, Q58C, Q58D, Q58E, Q59A, Q59B, Q59C, Q60A, Q60B, Q60C, Q60D, Q61A, Q61B, Q61C, Q62, Q63, Q64, Q65, Q66, Q67, Q68A, Q68B, Q69, Q70, Q70OTHER, Q71, Q72A, Q72B, Q73A, Q73B, Q74, Q75, Q76, Q77A, Q77B, Q77C, Q77D, Q78A, Q78B, Q79_CAM, Q80_CAM, Q81A_CAM, Q81B_CAM, Q81C_CAM, Q81D_CAM, Q82A_CAM, Q82B_CAM, Q82C_CAM, Q82D_CAM, Q84, Q84OTHER, Q85A, Q85B, Q86A, Q86B, Q86C, Q86D, Q87A, Q87B, Q87C, Q87D, Q88A,Q88B, Q88BOTHER, Q89A, Q89B, Q89C, Q89D, Q89E, Q89F, Q90, Q91A, Q91B, Q92A, Q92B, Q93, Q94, Q95A, Q95B, Q95C, Q96A, Q96B, Q97, Q98, Q98OTHER, Q99, Q99OTHER, Q100)

# Changing the types of all variables to factor, except the variable "age" which should be kept numeric
cameroonRound7 %>%
      mutate_each(funs(if(is.numeric(.)) as.factor(.) else .))

cameroonRound7$Q1 <- as.numeric(cameroonRound7$Q1)

# we need also to remove the variables that end with "OTHER" because most of the values (or responses) to the variables are of character or string type and generate the errors "NA/NaN/Inf" with the execution of some procedures such as the feature selection using "rfe"

cameroonRound7 <- select(cameroonRound7, - c(Q2AOTHER, Q2BOTHER, Q55PT3OTHER, Q55PT2OTHER, Q55PT1OTHER, Q70OTHER, Q98OTHER, Q84OTHER, Q88BOTHER, Q99OTHER))

# The outcome variable 1 is Q43A (Response to the question: "How much do you trust the President")
cameroonRound7$Q43A <- as.factor(cameroonRound7$Q43A)
str(cameroonRound7$Q43A)
summary(cameroonRound7$Q43A)
myTable <- table(cameroonRound7$Q43A)
prop.table(myTable)

# The outcome variable 2 is Q58A (approval of the way the president has performed his job over the last 12 month)
cameroonRound7$Q58A <- as.factor(cameroonRound7$Q58A)
str(cameroonRound7$Q58A)
summary(cameroonRound7$Q58A)
myTable <- table(cameroonRound7$Q58A)
prop.table(myTable)

# Crosstabulating approval and trust of the President
table(cameroonRound7$Q43A, cameroonRound7$Q58A)

# For the purpose of creating figure, we need to assign to Q43A ("Trust in President Paul Biya")and Q58A (Approval of the President's Job") their respective original values

cameroonRound7 <- mutate(cameroonRound7, trustInThePresident = ifelse(Q43A == 0, "1. Not at all", 
                                                                      ifelse(Q43A == 1, "2. Just a little",
                                                                             ifelse(Q43A == 2, "3. Somewhat", 
                                                                                    ifelse(Q43A == 3, "4. A lot",
                                                                                           ifelse(Q43A == 9, "5. Don't know", "5. Don't know"))))))

table(cameroonRound7$trustInThePresident)

cameroonRound7 <- mutate(cameroonRound7, approvalOfThePresident = ifelse(Q58A == 1, "1. Strongly disapprove", 
                                                                      ifelse(Q58A == 2, "2. Disapprove",
                                                                             ifelse(Q58A == 3, "3. Approve", 
                                                                                    ifelse(Q58A == 4, "4. Strongly approve",
                                                                                           ifelse(Q58A == 9, "5. Don't know", "5. Don't know"))))))

table(cameroonRound7$approvalOfThePresident)


# In order to simplify the interpretation of the results of the machine learning implementation, we need to create the variable "trust" with three categories of responses, by mutating Q43A and assigning the following value: 0 = "Distrust", 1 = "Trust", 2 = "Don't know"
cameroonRound7 <- mutate(cameroonRound7, trust = ifelse(Q43A == 0 | Q43A == 1, "2. Distrust",
                                           ifelse(Q43A == 2 | Q43A == 3, "1. Trust", "3. Don't know" )))
cameroonRound7$trust <- as.factor(cameroonRound7$trust)

# In order to simplify the interpretation of the results of the machine learning implementation, we need to create the variable "approval" with three categories of responses, by mutating Q58A and assigning the following value: 0 = "Disapprove", 1 = "Approve", 2 = "Don't know"
cameroonRound7 <- mutate(cameroonRound7, approval = ifelse(Q58A == 1 | Q58A == 2, "Disapprove",
                                           ifelse(Q58A == 3 | Q58A == 4, "Approve", "Don't know" )))
cameroonRound7$approval <- as.factor(cameroonRound7$approval)

table(cameroonRound7$trust, cameroonRound7$approval)

# In order to simplify the interpretation of the results of the machine learning implementation, we need to create the variable "voteForThePresident" with three categories of responses, by mutating Q99 and assigning the following value: 1 = "Vote for the President", 2 = "Vote for other candidates", 3 = "Don't know"
cameroonRound7 <- mutate(cameroonRound7, voteForThePresident = ifelse(Q99 == 1220, "1. Vote for the President",
                                           ifelse(Q99 == 1221 | Q99 == 1222 | Q99 == 1223 | Q99 == 1224, "2. Vote for other candidates", "3. Don't know" )))
cameroonRound7$voteForThePresident <- as.factor(cameroonRound7$voteForThePresident)
votingIntention <- table(cameroonRound7$voteForThePresident)

prop.table(votingIntention)

votingIntentionTrust <- table(cameroonRound7$trust, cameroonRound7$voteForThePresident)
prop.table(votingIntentionTrust)

votingIntentionApproval <- table(cameroonRound7$approval, cameroonRound7$voteForThePresident)
prop.table(votingIntentionApproval)


```


## Abstract

The durability (or stability) of the Paul Biya regime could be a blessing or a curse for the Cameroonian people. Indeed, almost every study on the relationship between regime durability (or political stability) and economic development would lead to the conclusion that the former is an important condition of the latter. Nevertheless, the fact that a country has been ruled for so many years by the same individual, without being a monarchy, is certainly a sign of autocracy. In addition, having the same leader for decades may also result in sclerosis and lack of innovations for further development of the country.

Most of those who attempted to explain the durability of the Biya regime adopted a top-down approach or  macro-level analysis. They tried to understand what the Biya regime did in order to stay in power for so long. In this sense, very few (if any) studies used a micro-level analysis or buttom-up approach, and attempted to understand the Cameroonian people's opinions and attitudes to explain why they were willing to stay under the same ruler for at least two generations. 

We adopt such micro-level analysis or butomm-up approach in this study. Using the data from round 6 and 7 of the Afrobarometer national surveys on Cameroon, we apply simple statistical analyses to find out to what extent the Cameroonian people trust their president and his regime, and to identify the factors that may influence them to support support him. As a result, we find that the great majority of the Cameroonian people not only trust their President Paul Biya, but they are also statisfied with the government performance, and believe that their country is democratic. In sum, using a micro-level analysis, we can draw the conclusion that the Biya regime has remained durable, because of the Cameroonian people's opinions and attitudes toward this regime.

**Keywords:** Africa, Cameroon, political stability, regime durability, micro-level analysis, Afrobarometer survey

## Introduction

President Paul Biya was elected for the seventh time as the president of Cameroon in October 2018.


The durability (or stability) of the Paul Biya regime could be a blessing or a curse for the Cameroonian people. Indeed, almost every study on the relationship between regime durability (or political stability) and economic development would lead to the conclusion that the former is an important condition of the latter. 

Nevertheless, the fact that a country has been ruled for so many years by the same individual, without being a monarchy, is certainly a sign of autocracy. In addition, having the same leader for decades may also result in sclerosis and lack of innovations for further development of the country.

[Purpose and reserach questions]

This study is based on a micro-level analysis or buttom-up approach to the durability of the Paul Biya regime. In this sense, instead of analyzing what the president and his regime are doing to stay in power for so long, we try to understand what the Cameroonian people think about their president and his rule. Thus, using the data from rounds 6 and 7 of the Afrobarometer national surveys on Cameroon, we try to find out to what extent the Cameroonian people trust their president and his regime and to identify the factors that may influence them in their support or lack thereof. 

## Literature Review

[Copied and pasted from the abstract: Most of those who attempted to explain the durability of the Biya regime adopted a top-down approach or  macro-level analysis. They tried to understand what the Biya regime did in order to stay in power for so long. In this sense, very few (if any) studies used a micro-level analysis or buttom-up approach, and attempted to understand the Cameroonian people's opinions and attitudes to explain why they were willing to stay under the same ruler for at least two generations.]





## Data and Methodology

Afrobarometer presents itself as "a pan-African, non-partisan research network that conducts public attitude surveys on democracy, governance, economic conditions, and related issues in more than 35 countries in Africa." (Afrobarometer, 2019) Cameroon was included in the rounds 5, 6, and 7 of the national surveys conducted by Afrobarometer. Even if they do not directly address the questions about the durability of the Paul Biya regime, the data from these surveys allow us to understand the Cameroonian people's opinions and attitudes toward the president and his regime, and explain why they are willing to stay under the same regime for so many years.


Proxy for support to the president and his regime = the trust in the president and the approval of the way the institutions handle their job (Q58)

Factors that may be associated with the trust in the institutions =

- regime performance
- satisfaction with the democracy
- satistfaction with freedom
- etc.


## Findings

### Descriptive Statistics

Afrobarometer's round 7 survey consists of the responses of 1204 respondents to some 297 questions. These respondents were randomly selected by Afrobarometer in taking into consideration the following factors:

- urban or rural location of the respondent,
- region of the respondent,
- age of the respondent,
- primary language spoken by the respondent at home

```{r}
# Distribution between urban and rural location of the respondents
table(cameroonRound7$URBRUR)

# Distribution of respondents by region of origin
table(cameroonRound7$REGION)

# Distribution of respondents by age
table(cameroonRound7$Q1)

# Distribution of respondents by primary language at home
table(cameroonRound7$Q2B)

```

In this study, we will focus on two major factors that may help us understand why Paul Biya's regime has been one of the most durable regime in Africa. These two factors are:

- the trust of the respondents in the President,
- the approval of the way the President performed his job

Assuming that the people who trust the President tend also to approve the way he performs his job, and tend also to vote for him or let him stay in power for a long time, we will identify in this study the factors that lead these people this trust and approval.

Concerning the level of trust, we find that:

- the great majority of Cameroonians (over 59%) trust "a lot" or "somewhat" the President (See Figure 1 below)
- about 35% of them do not trust "at all" or "trust a little"

```{r}
trustLevel <- table(cameroonRound7$trustInThePresident)

prop.table(trustLevel)

ggplot(data = cameroonRound7, aes(trustInThePresident)) +
  geom_bar()

```


Concerning the level of trust, we also find that:

```{r}
approvalLevel <- table(cameroonRound7$approvalOfThePresident)
prop.table(approvalLevel)

ggplot(data = cameroonRound7, aes(approvalOfThePresident)) +
  geom_bar()

```

- the great majority of Cameroonians (over 56%) either "approve"" or "strongly approve" the way the President performed his job, and  
- only about 33.5% of them do not "approve" or "stongly disapprove".



It is worth noting that the trust in the President and approval of his job performance are very low among the people in the Anglophone areas (North-East and South-East).

```{r}
# Subsetting the Anglophone respondents (inhabitants of the North-East and South_East regions, respectively No. 1228 and 1231 in the Afrobarometer survey)

anglophone <- filter(cameroonRound7, REGION == 1228 | REGION == 1231)
francophone <- filter(cameroonRound7, REGION != 1228 & REGION != 1231)

# Trust level among Anglophone
trustLevelAnglo <- table(anglophone$trustInThePresident)

prop.table(trustLevelAnglo)

ggplot(data = anglophone, aes(trustInThePresident)) +
  geom_bar()

# Approval level among Anglophone
approvalLevelAnglo <- table(anglophone$approvalOfThePresident)
prop.table(approvalLevelAnglo)

ggplot(data = anglophone, aes(approvalOfThePresident)) +
  geom_bar()

```
The correlation between "trust" and "approval":

In order to determine whether "trust" and "approval" are correlated and can be used interchangeable, we compute and plot the correlation between the two variables.

```{r}
# First, we need to transform the variables to numeric
cameroonRound7$Q43A <- as.numeric(cameroonRound7$Q43A)
cameroonRound7$Q58A <- as.numeric(cameroonRound7$Q58A)

cor(cameroonRound7$Q43A, cameroonRound7$Q58A)

```

The correlation between "trust", "approval", and "voting":

In order to determine whether "trust" or "approval" are correlated with "voting intention", we crosstabulate these variable Q99, which is the answer to the question "If presidential elections were held tomorrow, which party’s candidate would you vote for?"


```{r}

prop.table(votingIntention)

ggplot(data = cameroonRound7, aes(voteForThePresident)) +
  geom_bar()

votingIntentionTrust <- table(cameroonRound7$trust, cameroonRound7$voteForThePresident)
prop.table(votingIntentionTrust)

votingIntentionApproval <- table(cameroonRound7$approval, cameroonRound7$voteForThePresident)
prop.table(votingIntentionApproval)

```



### The Determinants of the Popular Trust in the President and Approval of his Job Performance

We run two machine learning scripts to identify the factors that led to the trust in the President and the approval of his job performance.

The results of these machine learning scripts are as follows:

1) The predictors of trust in the President are:




2) The predictors of the approval of the President's job performance are:



## Discussion and Conclusion




## References


